### 2025-09-10 docs/Homeownership_BankOfMomAndDad.tex
added short abstract
some shortening

\maketitle
\begin{abstract}
{\begin{spacing}{1.2}
Housing is the largest asset in U.S. household portfolios, and first-time homebuyers increasingly rely on parental transfers. This paper quantifies the contribution of parental transfers to the homeownership rate of young households. I build and estimate a life-cycle overlapping generations model with housing, where adult children and parents interact without commitment. I find that parental transfers account for 13 percentage points (27\%) of young households' homeownership. Transfers from wealthy parents not only help households overcome borrowing constraints, but also help sustain homeownership, mitigating the drawbacks of illiquidity. Surprisingly, policies lowering entry barriers to homeownership generally increase the reliance on parental wealth, whereas increased liquidity reduces it. Finally, I show that children of wealthy parents strategically use the illiquidity of housing as a commitment device to encourage transfers, resulting in a preference for illiquidity.
\end{spacing}}
\vspace{0.5cm}
\noindent \textbf{JEL:} D14, D15, E21, G51, R21

@@

\section{Data on Transfers, Family, and Housing}\label{sec:data}
I first present time trends in parental transfers for down payments, before describing the estimation sample taken from the PSID. Finally, I use the estimation sample to show that parental wealth is positively associated with better housing outcomes, both before and after purchase.

@@

\subsubsection{Descriptive Statistics: Who Receives Transfers?}
I now discuss descriptive statistics from the 2013 PSID sample for households aged 25 to 44 with an observed parent household. Table \ref{tab:descrstats} contains the means of variables, by age, wealth, homeownership, and transfer receipt. Though limited to households with observed parents, the sample appears reasonably representative. Average income by age closely matches estimates from the SCF \cite[see e.g.,][]{kuhn20162013}, though average wealth is slightly lower.

There are several main takeaways from the subgroup analysis. In 2012, 21\% of young households received a transfer, and transfers averaged \$3,960. Receivers have significantly richer parents, have similar wealth and income as nonreceivers, and are less likely to own. Receivers are more likely to transition from renting to owning, especially in the age groups where households are most likely to buy. The key determinant of transfer receipt seems to be parental wealth.

@@

\section{A Quantitative Model of Parental Transfers and Homeownership}\label{sec:model}
This section describes my life-cycle model of housing choices with overlapping generations, idiosyncratic earnings risk, and altruistic inter-vivos transfers from parents to their offspring. To answer how important parental transfers are for housing outcomes and how transfers interact with popular housing policies, I combine two models: altruism without commitment and housing.

@@

\textit{Period Length:} Each period corresponds to two years, matching the PSID interview frequency. I report all parameters in biennial form (e.g., $\beta$ measures how much households discount two years).

@@

To calibrate the stochastic income process $y_{i,a}$, I first set $N_y=3$ to create a three-state Markov chain. The sample is divided into age-specific income tertiles. Within each age, I set $v_{i,a}$ as the ratio of the median income for that tertile to the overall median income (Figure \ref{fig:y_ia}). The empirical age-dependent transition matrices provide the transition matrix (Figure \ref{fig:Piv}).

@@

\textit{Initial Conditions of the Young $y_{c,25},x_{c,25}\sim F(x_{c,53},y_{c,53})$:} 
The initial distribution is set to match the intergenerational correlation in wealth and income. I estimate it nonparametrically for transparency. First, I limit the sample to households aged 24 to 26 with parents 16 to 40 years older. I then divide wealth and income into four and three quantiles, respectively, for both parents and children. The PDF $F(\cdot)$ is then given by the empirical probability of each parent-child combination. The wealth level of a quartile is defined by its median value (Fig. \ref{fig:inidistr})

@@

I assume that housing supply ($H^S$) is log-linear in the house price:
\begin{equation}
\label{eq:hsupply}
\ln H^S = \alpha_0 + \alpha_1 \ln p,
\end{equation}
where $\alpha_1$ is the aggregate elasticity of supply to prices. I use a closed-form supply function because the model is too complex to introduce a construction sector. The rent-to-price ratio is unchanged, as in \cite{Kaplan2020}, where rental units are supplied by deep-pocketed landlords who convert rentals to owner-occupied units if the price deviates from the present value of rents. Letting $\alpha_1\to\infty$ yields a perfectly elastic supply function, leaving prices constant, as in the previous section. I first set the elasticity to 3, a typical estimated value for the United States \cite[see e.g.,][]{saiz2010geographic,aastveit2023changing}. Finally, I set the elasticity to 1 as a reasonable lower bound for the long-run aggregate elasticity.

@@

\textit{Tighter LTVs:} LTV caps, a standard macroprudential tool, directly regulate down-payment requirements; as shown in Figure~\ref{fig:motivation}, reliance on down-payment assistance moves with changes in mortgage regulation. I tighten the LTV limit by 3 percentage points. Homeownership among the young falls by 7 percentage points (15\%), with larger declines for children of middle-wealth parents, flattening the parental wealth gradient. The mechanism is that these families rely on modest transfers to clear borrowing constraints; poorer households remain constrained, and richer ones are largely unaffected. Symmetrically, loosening borrowing constraints would raise homeownership and amplify the role of parental wealth. Indeed, \citet{wold2024housing} find that owners with poorer parents have lower LTV ratios at origination.

\textit{Lowering PMI:} Several government-backed mortgages (e.g., VA, USDA) do not require PMI; to mimic easing on this margin, I relax the PMI limit by 7 percentage points. Homeownership rises by 1 percentage point (2\%). As with the LTV rule, gains are larger for households with parents in the middle of the wealth distribution, further widening the parental-wealth gradient. Thus, relaxing PMI increases both overall homeownership and the importance of parental wealth.

\textit{Lowering House Prices:} Although there is no single “price lever,” policy shapes prices through supply regulation, taxes, interest rates, and subsidies. I capture these channels in reduced form by lowering house prices by 2\%. Homeownership increases for all households, but more so among those with the richest parents, thereby increasing the role of parental wealth.

\textit{Reducing Purchase Costs:}
Closing-cost programs and first-time-buyer relief---e.g., England’s 2017 stamp duty cut for first-time buyers---directly lower entry costs; to study this margin’s interaction with parental transfers, I lower the purchase cost ($m_b$) by 2 percentage points, equivalent to a grant of about \$6,000. While homeownership rises, the gains are concentrated among households with wealthy parents, amplifying the parental wealth gradient. Lowering purchase costs has little effect for renters with non-wealthy parents, who remain constrained by LTV limits and sales costs. In contrast, for households with wealthy parents, lower purchase costs reduce the effective price—the main remaining barrier to ownership. This suggests that grants or credits for first-time buyers primarily benefit households with richer parents.

@@

\appendix

@@

\section{Transfers and the Rent-to-Own Transition}\label{app:rent_to_own}
A large literature documents that parental transfers increase the probability that renters become homeowners (e.g., \citealp{wold2024housing}; \citealp{Blickle2019}; \citealp{benetton2022dynastic}; \citealp{Guiso2002}; \citealp{Engelhardt1998}). I replicate this result to highlight the importance of transfers for housing outcomes.

I estimate
\begin{equation}	\label{eq:transferrent}
	\Pr\left(Own_{i,t+2}=1 \mid Own_{i,t}=0\right)
	= \alpha + \beta\,Transfer_{i,t+2}
	  + \gamma' X_{i,t} + \varepsilon_{i,t},
\end{equation}
where \(Own_{i,t}=0\) denotes renting at time \(t\), \(Transfer_{i,t+2}=1\) if the household received a transfer between survey waves, and \(X_{i,t}\) includes log parental wealth, log household income, log household net worth, log household size, age, parental age, and dummies for  education, marital status, race, household composition changes, and state and year fixed effects. The sample is limited to renting households aged 25 to 44 who have never been observed as homeowners. Table \ref{tab:newowners} reports the estimated coefficients. These regressions are similar to those in \cite{Lee2018}, who also use the PSID.  It is important to note that transfer receipt is not exogenous---many transfers may be made precisely because the renter is buying a home.

@@

\begin{figure}
    \caption{Calibration of Parental Income Shocks}\label{fig:nu}
	{\centering
    \includegraphics[width=0.5\textwidth]{../tabfig/empirical/lifecycleproductivity_old_3}
    \par}    {\begin{footnotesize} \textit{Notes:}
        Dashed lines are the empirical medians, and solid lines are fitted second-order polynomials used in model calibration. The lines denote the value of $\nu_{i,a}$ in the first tertile (black), middle (gray), and top (blue) by age.
    \end{footnotesize}}
\end{figure}

@@

\begin{figure}
	\caption{Event Study at Various Cutoffs: Housing Consumption}\label{fig:housinggrowthrates_threshold} 
	\noindent%
	\includegraphics[width=0.5\textwidth]{../tabfig/descr/PSID_housinggrowthpoorall_both}%
	\includegraphics[width=0.5\textwidth]{../tabfig/descr/PSID_housinggrowthrichall_both}

@@

I use the same estimation samples and variable definitions. I then run linear regressions where I interact years relative to unemployment with a dummy for having wealthy parents at unemployment. The set of controls includes dummies for children's wealth quintiles, a full set of age, year, and state dummies, and dummy variables for college, high school, marriage, race, log income and log family size. Standard errors are clustered at the household level. The results are plotted in Figure \ref{fig:housinggrowthrates_controls}. The main result is that households with wealthier parents do not significantly decrease housing consumption at unemployment, while those with poor parents, on average, decrease housing consumption by 8 percent ($p<0.05$). However, we cannot reject the hypothesis that the effect of unemployment is the same for both groups.
\begin{figure}
	\caption{Event Study with Controls: Housing Consumption}\label{fig:housinggrowthrates_controls}
	\includegraphics[width=0.5\textwidth]{../tabfig/descr/PSID_housinggrowthpoor_both_controls}%
	\includegraphics[width=0.5\textwidth]{../tabfig/descr/PSID_housinggrowthrich_both_controls}

>>>>

\maketitle
\begin{abstract}
{\begin{spacing}{1.2}
Housing is the largest asset in U.S. household portfolios, and first-time homebuyers increasingly rely on parental transfers. This paper quantifies the contribution of parental transfers to the homeownership rate of young households. I build and estimate a life-cycle overlapping generations model with housing, where adult children and parents interact without commitment. I find that parental transfers account for 13 percentage points (27\%) of young households' homeownership. Transfers from wealthy parents not only help households overcome borrowing constraints, but also help sustain homeownership, mitigating the drawbacks of illiquidity. Surprisingly, policies lowering entry barriers to homeownership generally increase the reliance on parental wealth, whereas increased liquidity reduces it. Finally, I show that children of wealthy parents strategically use the illiquidity of housing as a commitment device to encourage transfers, resulting in a preference for illiquidity.

% <=100 word abstract This paper quantifies how parental transfers shape young adults’ homeownership. I estimate a life-cycle OLG model with housing in which parents and children interact without commitment. Parental transfers account for 13 percentage points (27\%) of young households’ homeownership. Transfers from wealthy parents both relax borrowing constraints and sustain ownership by mitigating housing illiquidity. Policy experiments show that easing entry barriers increases reliance on parental wealth, whereas increasing housing liquidity reduces it. Finally, I show that children of wealthy parents strategically use the illiquidity of housing as a commitment device to encourage transfers, resulting in a preference for illiquidity.
\end{spacing}}
\vspace{0.5cm}
\noindent \textbf{JEL:} D14, D15, E21, G51, R21

@@

\section{Data on Transfers, Family, and Housing}\label{sec:data}
I first present time trends in parental transfers for down payments, before describing the estimation sample. Finally, I use the estimation sample to show that parental wealth is positively associated with better housing outcomes, both before and after purchase.

@@

\subsubsection{Descriptive Statistics: Who Receives Transfers?}
I now discuss descriptive statistics from the 2013 PSID sample for households aged 25 to 44 with an observed parent household. Table \ref{tab:descrstats} contains the means of variables, by age, wealth, homeownership, and transfer receipt. Though limited to households with observed parents, the sample appears representative. Average income by age closely matches estimates from the SCF \cite[see e.g.,][]{kuhn20162013}, though average wealth is slightly lower.

In 2012, 21\% of young households received a transfer, and transfers averaged \$3,960. Receivers have significantly richer parents, have similar wealth and income as nonreceivers, and are less likely to own. Receivers are more likely to transition from renting to owning, especially in the age groups where households are most likely to buy. The key determinant of transfer receipt seems to be parental wealth.

@@

\section{A Model of Transfers and Homeownership}\label{sec:model}
This section describes my life-cycle model of housing choices with overlapping generations, idiosyncratic earnings risk, and altruistic inter-vivos transfers from parents to their offspring. To answer how important parental transfers are for housing outcomes and how transfers interact with popular housing policies, I combine two models: altruism without commitment and housing.

@@

\textit{Period Length:} Each period corresponds to two years, matching the PSID interview frequency. I report all parameters in biennial form (e.g., $\beta$ discounts two years).

@@

To calibrate the stochastic income process $y_{i,a}$, I first set $N_y=3$ to create a three-state Markov chain. The sample is divided into age-specific income tertiles. Within each age, I set $v_{i,a}$ as the ratio of the median income for that tertile to the overall median income (Figure \ref{fig:y_ia}). The age-dependent transition matrices are taken from the data (Figure \ref{fig:Piv}).

@@

\textit{Initial Conditions of the Young $y_{c,25},x_{c,25}\sim F(x_{c,53},y_{c,53})$:} 
The initial distribution is set to match the intergenerational correlation in wealth and income, estimated nonparametrically for transparency. First, I limit the sample to households aged 24 to 26 with parents 16 to 40 years older. I then divide wealth and income into four and three quantiles, respectively, for both parents and children. The PDF $F(\cdot)$ is then given by the empirical probability of each parent-child combination. The wealth level of a quartile is its median value (Fig. \ref{fig:inidistr})

@@

I assume that housing supply ($H^S$) is log-linear in the house price:
\begin{equation}
\label{eq:hsupply}
\ln H^S = \alpha_0 + \alpha_1 \ln p,
\end{equation}
where $\alpha_1$ is the aggregate elasticity of supply to prices. I use a closed-form supply function because the model is too complex to introduce a construction sector. The rent-to-price ratio is unchanged, as in \cite{Kaplan2020}, where rental units are supplied by deep-pocketed landlords who convert them to owner-occupied units if the price deviates from the present value of rents. Letting $\alpha_1\to\infty$ yields a perfectly elastic supply function, leaving prices constant, as in the previous section. I experiment with setting the elasticity to 3, a typical estimated value for the United States \cite[see e.g.,][]{saiz2010geographic,aastveit2023changing}and to 1, a reasonable lower bound for the long-run aggregate elasticity.

@@

\textit{Tighter LTVs:} LTV regulation is a standard macroprudential tool and as shown in Figure~\ref{fig:motivation} reliance on down-payment assistance moves with changes in mortgage regulation. I tighten the LTV limit by 3 percentage points. Homeownership among the young falls by 7 percentage points (15\%), with larger declines for children of middle-wealth parents, flattening the parental wealth gradient. The mechanism is that these families rely on modest transfers to clear borrowing constraints; poorer households remain constrained, and richer ones are largely unaffected. Symmetrically, loosening borrowing constraints would raise homeownership and amplify the role of parental wealth. Indeed, \citet{wold2024housing} find that owners with poorer parents have lower LTV ratios at origination.

\textit{Lowering PMI:} Several government-backed mortgages (e.g., VA, USDA) do not require PMI; to mimic easing on this margin, I relax the PMI limit by 7 percentage points. Homeownership rises by 1 percentage point (2\%). As with the LTV, gains are larger for households with parents in the middle of the wealth distribution, further widening the parental-wealth gradient. Thus, relaxing PMI increases both overall homeownership and the importance of parental wealth.

\textit{Lowering House Prices:} While policy does not set house prices, prices are affected by supply regulation, taxes, interest rates, and subsidies. I capture these channels in reduced form by lowering house prices by 2\%. Homeownership increases for all households, but more so among those with the richest parents, thereby increasing the role of parental wealth.

\textit{Reducing Purchase Costs:}
Closing-cost programs and first-time-buyer relief---e.g., England’s 2017 stamp duty cut for first-time buyers---directly lower entry costs. I lower the purchase cost ($m_b$) by 2 percentage points, equivalent to a grant of about \$6,000. While homeownership rises, the gains are concentrated among households with wealthy parents, amplifying the parental wealth gradient. Lowering purchase costs has little effect for renters with non-wealthy parents, who remain constrained by LTV limits and sales costs. In contrast, for households with wealthy parents, lower purchase costs reduce the effective price—the main remaining barrier to ownership. This suggests that grants or credits for first-time buyers primarily benefit households with richer parents.

@@

% \cleardoublepage  % This ensures you start on a new page
% \thispagestyle{empty}  % This removes page number and headers/footers
% \vspace*{\fill}  % This pushes the text to the vertical center of the page
% \begin{center}
% \Large\textbf{Online Appendix}  % This creates the text, centered and bold
% \end{center}
% \vspace*{\fill}  % This completes the vertical centering
% \clearpage  % This ends the current page
\appendix

@@

\section{Transfers and the Rent-to-Own Transition}\label{app:rent_to_own}
A large literature documents that parental transfers increase the probability that renters become homeowners (e.g., \citealp{wold2024housing}; \citealp{Blickle2019}; \citealp{benetton2022dynastic}; \citealp{Guiso2002}; \citealp{Engelhardt1998}). I replicate this result to highlight the importance of transfers for housing outcomes, by estimating
\begin{equation}	\label{eq:transferrent}
	\Pr\left(Own_{i,t+2}=1 \mid Own_{i,t}=0\right)
	= \alpha + \beta\,Transfer_{i,t+2}
	  + \gamma' X_{i,t} + \varepsilon_{i,t},
\end{equation}
where \(Own_{i,t}=0\) denotes renting at time \(t\), \(Transfer_{i,t+2}=1\) if the household received a transfer between survey waves, and \(X_{i,t}\) includes log parental wealth, log household income, log household net worth, log household size, age, parental age, and dummies for  education, marital status, race, household composition changes, and state and year fixed effects. The sample is limited to renting households aged 25 to 44 who have never been observed as homeowners. Table \ref{tab:newowners} reports the estimated coefficients. These regressions are similar to those in \cite{Lee2018}, who also use the PSID.  It is important to note that transfer receipt is not exogenous---many transfers may be made precisely because the renter is buying a home.

@@

\begin{figure}
    \caption{Calibration of Parental Income Shocks}\label{fig:nu}
	{\centering
    \includegraphics[width=0.45\textwidth]{../tabfig/empirical/lifecycleproductivity_old_3}
    \par}    {\begin{footnotesize} \textit{Notes:}
        Dashed lines are the empirical medians, and solid lines are fitted second-order polynomials used in model calibration. The lines denote the value of $\nu_{i,a}$ in the first tertile (black), middle (gray), and top (blue) by age.
    \end{footnotesize}}
\end{figure}

@@

\begin{figure}
	\caption{Event Study at Various Cutoffs: Housing Consumption}\label{fig:housinggrowthrates_threshold} 
	\noindent%
	\includegraphics[width=0.45\textwidth]{../tabfig/descr/PSID_housinggrowthpoorall_both}%
	\includegraphics[width=0.45\textwidth]{../tabfig/descr/PSID_housinggrowthrichall_both}

@@

I use the same estimation samples and variable definitions. I then run linear regressions where I interact years relative to unemployment with a dummy for having wealthy parents at unemployment. The set of controls includes dummies for children's wealth quintiles, a full set of age, year, and state dummies, and dummy variables for college, high school, marriage, race, log income and log family size. Standard errors are clustered at the household level. The results are plotted in Figure \ref{fig:housinggrowthrates_controls}. The main result is that households with wealthier parents do not significantly decrease housing consumption at unemployment, while those with poor parents, on average, decrease housing consumption by 8 percent ($p<0.05$). However, we cannot reject the hypothesis that the effect of unemployment is the same for both groups.
\begin{figure}
	\caption{Event Study with Controls: Housing Consumption}\label{fig:housinggrowthrates_controls}
	\includegraphics[width=0.45\textwidth]{../tabfig/descr/PSID_housinggrowthpoor_both_controls}%
	\includegraphics[width=0.45\textwidth]{../tabfig/descr/PSID_housinggrowthrich_both_controls}
